---
title: "R Practice: Tidy Data and Joins"
format: html
date: 03/12/2026
author: Lala Movsesian
editor: 
  markdown: 
    wrap: sentence
---

loading libraries:

```{r}
library(readr)
library(dplyr)
library(here)
library(lubridate)
```

## Introduction

This is a data wrangling practice following the steps described in the course textbook week 7.

## Part 1 Data

This is data from the EDI Data Portal Ecological and social interactions in urban parks: bird surveys in local parks in the central Arizona-Phoenix metropolitan area.

```{r}
bird_observations <- read.csv(here::here("data","raw","birds.csv"))
sites <- read.csv(here::here("data","raw","sites.csv"))
surveys <- read.csv(here::here("data","raw","surveys.csv"))
taxalist <- read.csv(here::here("data","raw","taxalist.csv"))
```

## Part 2

uncomment to explore the data:

```{r}
#head(bird_observations)
#head(sites)
#head(surveys)
#head(taxalist)
```

## Part 3

making a subset to show only species brown-headed cowbird and red-winged blackbird, and sites LI-W and NU-C

```{r}
birds_subset <- bird_observations %>% 
  filter(species_id %in% c("BHCO", "RWBL")) %>% 
  filter(site_id %in% c("LI-W", "NU-C"))
```

## Part 4

### Question 4a:

what do you expect the outcome data frame when doing left_join() between birds_subset and sites to look like?
What observations do you expect in the outcome data frame?

**Answer:** row count same as birds_subset, added columns from sites (park code, name, district, ...)

Adding site information from sites data using left joint and site_id as key:

```{r}

birds_left <- birds_subset %>% 
  left_join(y = sites, by = "site_id")

```

## Part 5

### Question 5a:

What do you expect a full_join() between birds_subset and sites to contain?
Write this in your Quarto Document or tell a neighbor.

**Answer:** adding all sites, so a lot of rows where most don't have any observation/bird data in birds_subset.
This is not a reasonable approach in this case.

## Part 6

### Question 6a:

What do you expect an inner_join() between birds_subset and taxalist to contain?

**Answer:** Using species_id as key, all rows from bird_subset will be in the new df since both of the unique values in species_id from bird_subset exist in taxalist.
Columns with species common name and asu_itis will be added to the new df.

### Question 6b:

What would you get if instead of inner_join() you’d used left_join() for this example?
Write code for the left join and check.

**Answer:** same as inner_joint, since all unique values of the key in bird_subset exist in taxalist

### Question 6c: Why does that make sense for this scenario?

In what case would you expect the outcome to differ from an inner_join()?
Write this in your Quarto Document or tell a neighbor.

**Answer:** if there are species_id values in bird_subset that are not in taxalist

Adding taxa information from taxalist data using inner joint and species_id as key

```{r}
birds_inner <- birds_subset %>% 
  inner_join(y = taxalist, by = "species_id")
```

## Part 7

Renaming notes column in bird_obserrvation, so it differs from notes column in survey:

```{r}
bird_observations <- bird_observations %>% 
  rename(bird_obs_notes = notes)
```

Creating a df with all taxa, site, and survey information (merging all 4 datasets), only showing survey_date, common_name, park_name, bird_count and observer name:

```{r}
bird_obs_subset <- bird_observations %>% 
  full_join(y = taxalist, by = "species_id") %>% 
  full_join(y = sites, by = "site_id") %>% 
  full_join(y = surveys, by = c("site_id", "survey_id")) %>%  
  select(survey_date, common_name, park_name, bird_count, observer)
```

## Part 8

Checking observer names:

```{r}
unique(bird_obs_subset$observer)
```

Changing Josh Burns to J.
Burns since *someone forgot to keep their data entry consistent* :)) then double checking:

```{r}
bird_obs_subset <- bird_obs_subset %>% 
  mutate(observer = if_else(condition = observer == "Josh Burns", 
                            true = "J. Burns", 
                            false = observer))

unique(bird_obs_subset$observer)
```

## Part 9

Adding new column to show month of survey:

```{r}
bird_obs_subset$survey_month <- month(ymd_hms(bird_obs_subset$survey_date))
```

Moving month column to right after survey date:

```{r}
bird_obs_subset <- bird_obs_subset %>% 
  relocate(survey_month, .after = survey_date)
```

Filtering for some parks (Lindo, Orme, Palomino, Sonrisa):

```{r}
bird_obs_subset_parks <- bird_obs_subset %>% 
  filter(park_name %in% c("Lindo", "Orme", "Palomino", "Sonrisa"))
```

quick check:

```{r}
unique(bird_obs_subset_parks$park_name)
unique(bird_obs_subset_parks$survey_month)
unique(bird_obs_subset_parks$bird_count)
```

There are NAs in bird_count and survey_month, excluding those rows, and calculating total number of birds per month per park

```{r}
bird_month_counts <- bird_obs_subset_parks %>%
  filter(!is.na(bird_count), !is.na(survey_month)) %>%
  group_by(park_name, survey_month) %>%
  summarise(total_birds = sum(bird_count))
```

## Summary:

### Conclusion

Final data shows a difference in bird count in different parks with Lindo mostly having the highest count.
There seems to be a difference in months across the years, spring/summer months show an overall higher count compared to fall/winter.

### Questions

Does it give you the outcome you expected?
\
Not at first, there were rows of NA date/count, was resolved by checking main dfs for NA values and excluding NA counts.

Is it informative?
\
It can be used to test seasonal variability in the number of birds.
Very simplified since year and species (biodiversity) are not taken into consideration.

How would you improve this wrangling process?
\
Check for NAs and other "problematic" data before joins.
